Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-08-21 paper/main.tex
whitespace and fix some errors

\input{paper/defs.tex}

@@

\begin{figure*}[th]
\begin{center}
\includegraphics[width=\textwidth]{paper/proper-motion.pdf}
\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
\label{fig:pm}
\end{center}
\end{figure*}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{paper/tracks.pdf}
\caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
\textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
\textbf{Upper right}: Distance as a function of Pal 5 longitude, $\phi_1$, for the RRLs. The shaded (orange) band shows the 16--84th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
\textbf{Lower panels}: Same as upper right, but for the proper motion in right ascension, $\mu_\alpha$, and declination, $\mu_\delta$.
}
\label{fig:trackmembers}
\end{center}
\end{figure*}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{paper/members.pdf}
\caption{Sky positions of RRL...todo}
\label{fig:members}
\end{center}
\end{figure*}

@@

We find 15 RRLs in the cluster tails, $\sim11$ of \type~ab and $\sim4$ of \type~c. Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5. Figure~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5. The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistently with their mean period of 0.565~d(XX).  Our findings of the updated ratio of \rrab~to~\rrc, mean period and Period-Amplitude distribution of the \rrab, all support an OoI classification for Pal~5, consistent with the cluster's metallicity.

Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars. This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\Feh>-1.5$][]{Monelli2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy. \todo{CM:( any other point to be made about this?)}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{paper/rrls_PA.pdf}
\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}\label{fig:PA_diagram}
\end{center}
\end{figure}

@@

To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab RRL, based on data from Galactic dwarf galaxies and globular clusters. Here we use it including the data for the \typec~stars.
Correcting the observed number of 25~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$.

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{paper/ps1_completeness.pdf}
% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
% \end{center}
% \end{figure}

\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\Feh$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}

>>>>

\input{defs.tex}
\graphicspath{paper/}

@@

\begin{figure*}[th]
\begin{center}
\includegraphics[width=\textwidth]{proper-motion.pdf}
\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
\label{fig:pm}
\end{center}
\end{figure*}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{tracks.pdf}
\caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
\textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
\textbf{Upper right}: Distance as a function of Pal 5 longitude, $\phi_1$, for the RRLs. The shaded (orange) band shows the 16--84th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
\textbf{Lower panels}: Same as upper right, but for the proper motion in right ascension, $\mu_\alpha$, and declination, $\mu_\delta$.
}
\label{fig:trackmembers}
\end{center}
\end{figure*}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{members.pdf}
\caption{Sky positions of RRL...todo}
\label{fig:members}
\end{center}
\end{figure*}

@@

We find 15 RRLs in the cluster tails, $\sim11$ of \typeab\ and $\sim4$ of \typec. Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5. Figure~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5. The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistently with their mean period of 0.565~d(XX).  Our findings of the updated ratio of \rrab~to~\rrc, mean period and Period-Amplitude distribution of the \rrab, all support an OoI classification for Pal~5, consistent with the cluster's metallicity.

Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars. This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy. \todo{CM:( any other point to be made about this?)}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{rrls_PA.pdf}
\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}\label{fig:PA_diagram}
\end{center}
\end{figure}

@@

To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters. Here we use it including the data for the \typec~stars.
Correcting the observed number of 25~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$.

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
% \end{center}
% \end{figure}

\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
